---
title: "Practical Report 9"
output:
  html_document:
    df_print: paged
    code_folding: hide
  bookdown::gitbook:
    highlight: monochrome
    code_folding: hide
    global_numbering: yes
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    highlight: monochrome
fontsize: 12pt
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
require(knitr)
pdf_doc <- knitr::is_latex_output()
if (pdf_doc) {opts_chunk$set(echo=TRUE, fig.align="center", out.width="60%", comment="")} else {opts_chunk$set(comment="")}
```

# Effectiveness of blood pressure treatment

We begin with the model from Practical 7, that is
$$
\begin{align*}
{\mathbb E}[\texttt{BPchange}_i]\quad = \quad \beta_0+\beta_1\texttt{BPinit}_i+\beta_2\texttt{Dose}_i, \qquad \mbox{where } i=1, \ldots, 40.
\end{align*}
$$

```{r, LoadData}
bpdata <- read.csv("../data/bpdata.csv")
lm.out1 <- lm(BPchange ~ BPinit + Dose, data=bpdata)
```

## Part a

We perform the test by `anova` between null model and `lm.out1`.

```{r, GlobalFtest}
# Fit null model
lm.out0 <- lm(BPchange ~ 1, data=bpdata)
# Use the anova function with two arguments
anova(lm.out0, lm.out1)
```

The table we want is

| Source | d.f. | SS | MS | F |
|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
| Regression | 2 | $\sum_{i=1}^{n} (\hat{y}_i - \bar{y})^2$ | $\frac{\text{RegrSS}}{p-1}$ | 96.128 |
| Residual | 37 | $\sum_{i=1}^{n} (y_i - \hat{y}_i)^2$ | $\frac{\text{ResidSS}}{n-p}$ |  |
| Total | 39 | $\sum_{i=1}^{n} (y_i - \bar{y})^2$ |  |  |

The output gives the observed $F$-statistic and $p$-value but we still
need to compute the mean squares and sum of squares to construct the
full ANOVA table for the existence of regression.

From the table, we know directly that residual sum of squares (ResidSS)
is $$
\sum_{i=1}^n (y_i - \hat{y}_i)^2 = 42.248,
$$ and the regression sum of squares (RegrSS) is $$
\begin{align*}
\sum_{i=1}^n (\hat{y}_i - \bar{y})^2 &= \sum_{i=1}^n (y_i - \bar{y})^2 - \sum_{i=1}^n (y_i - \hat{y}_i)^2\\
&= \text{ResidSS of Model 1} - \text{ResidSS of Model 2}\\
&= 261.775 - 42.248 = 219.527
\end{align*}
$$

For the mean squares, we have

-   ${\rm RegrMS}  = \frac{{\rm RegrSS}}{{\rm df(RegrSS)}} = \frac{219.527}{2} = 109.762,$

-   ${\rm ResidMS}  = \frac{{\rm ResidSS}}{{\rm df(ResidSS)}} = \frac{42.248}{37} = 1.142,$

Recall that the test for existence of regression is based on $$
\begin{align*}
H_0:\:  \beta_1=\beta_2=0 \qquad \text{vs} \qquad 
H_1:\:  \mbox{ at least one of } \beta_1,\beta_2 \mbox{ is non-zero}.
\end{align*}
$$ We have $n=40$ and $p=3$ from data and model parameters. Thus, under
$H_0$, $F \sim F_{p-1,n-p} = F_{2, 37}$. We can calculate the p-value
given by $P(F_{2, 37} \geq F_{\rm obs})$ where $F_{\rm obs}=96.128$ and
obtain
`r formatC(pf(96.128, df1=2, df2=37, lower.tail=FALSE), format = "e", digits = 2)`
which is substantially smaller than the significance level of 5%. Thus
we have very strong evidence to reject $H_0$. We conclude that at least
one of the two explanatory variables, that is the initial blood pressure
(`BPinit`) and the dosage of the new drug (`Dose`), are informative
about the change in blood pressure (`BPchange`).

## Part b

The `anova()` function with one argument produces a sequential ANOVA
which shows what happens when, starting from the null model (with only
the intercept), we add the terms of the model sequentially from left to
right. If the formula is $$
\begin{align*}
{\tt BPchange \quad \sim  \quad BPinit + Dose}
\end{align*}
$$ when the `R` output shows the effect of adding `BPinit` then `Dose`.

```{r, SequentialAnova}
anova(lm.out1)
```

## Part c

By controlling for initial blood pressure, this question is asking the
effect of new drug on blood pressure. This is the same as asking about
the significance of the parameter for the `Dose` term. Hence, we can
apply a hypothesis test as $$
\begin{align*}
    H_0:\:  \beta_2 = 0  \quad \text{versus} \quad
    H_1:\:  \beta_2 \neq 0.
\end{align*}
$$ As `Dose` was entered second in the model, this test is reported on
in the sequential ANOVA in the row corresponding to `Dose`. As the
$p$-value is less than 0.001, there is very strong evidence to reject
$H_0$. We thus conclude that, controlling for initial blood pressure,
the new drug has an effect on blood pressure.

We can also use a $t$-test which is reported on in the summary output of
the model. In fact, this two-sided $t$-test is equivalent to the partial
$F$-test reported on in the sequential ANOVA. The observed $T$-statistic
is equal to `-12.284` with a $p$-value of `1.27e-14`. (Note that the
square of the observed $T$-statistic is equal to the observed
$F$-statistic and they both have the same $p$-value.)

```{r summary}
summary(lm.out1)
```

## Part d

For the degrees of freedom column, we get:

-   The total degrees of freedom is the sum of all 3 entries in the
    column labelled "Df":\
    $(39 = 1 + 1 + 37)$

-   The regression degrees of freedom is the sum of the two rows for the
    terms `Dose` and `BPinit`: $(2 = 1 + 1)$

-   The residual degrees of freedom is from the third row labelled
    \`\`Residuals'': $37$.

For the sum of squares column, we get:

-   The total sum of squares is the sum of all 3 entries in the column labelled \`\`Sum Sq": $(261.775 = 47.214 + 172.313 + 42.248).$

-   The regression sum of squares is the sum of the two rows for the terms `Dose` and `BPinit`:
$(219.527 = 47.214 + 172.313).$

-   The residual sum of squares is from the third row labelled \`\`Residuals" (42.248.)

Hence, the $F$-statistic can be computed as

-   ${\rm RegrMS}  = \frac{{\rm RegrSS}}{{\rm df(RegrSS)}} = \frac{219.527}{2} = 109.762,$

-   ${\rm ResidMS}  = \frac{{\rm ResidSS}}{{\rm df(ResidSS)}} = \frac{42.248}{37} = 1.142,$

-   ${F}_{\rm obs} = \frac{\rm RegrMS}{\rm ResidMS} = \frac{109.762}{1.142} = 96.128.$

# Model selection statistics

In Experiment 1 we generate data for 50 predictor variables
$X_1, \ldots, X_{50}$ all of which are non-informative about the
response $Y$ as the latter is sampled independently from the predictor
variables. We then consider a sequence of models, starting with a simple
linear regression of $Y$ on $X_1$ and adding the other predictor
variables one at a time. For each model the coefficient of determination
$R^2$ is computed. The plot below shows the resulting sequence of $R^2$
values.

```{r, echo=T, fig.cap="Figure 1: A graphical illustration of the coefficient of determination as the number of non-informative predictor variables increases in the model."}
# Generate data
set.seed(22)
x <- matrix(rnorm(5000),100,50)
y <- rnorm(100)
data <- as.data.frame(cbind(y,x))
# Fit linear regression models with increasing number of variables
r2 <- rep(NA,50)
for (j in 1:50){
  data_temp <- data[ ,1:(j+1)]
  r2[j] <- summary(lm(y ~ ., data=data_temp))$r.squared 
}
# Plot the sequence of R^2 statistics
plot(r2,type="l",lwd=1.5,col="blue",
     xlab="number of predictor variables",ylab=expression(R^2))
points(r2,col="blue",pch=20)
```

We observe that despite the predictor variables all being
non-informative about the response, the coefficient of determination
$R^2$ continues to increase as we add more predictor variables to the
model.

In Experiment 2 we generate data for 10 predictor variables. The
response is sampled from the model $$
\begin{align*}
Y_i = 1 + 0.25x_{i,1}  + 0.5x_{i,2} + 0.5 x_{i,3}  + \epsilon_i, \qquad \mbox{where }  \epsilon_i \sim {\cal N}(0,1) \;  {\rm iid},  \qquad i=1, \ldots, 10.
\end{align*}
$$ So only the first three variables are informative about $Y$, while
the other predictor variables are not.

```{r, echo=T, fig.cap="Figure 2: A graphical illustration of the difference of behaviour between the coefficient of determination and the adjusted coefficient of determination as the number of predictor variables increases in the model."}
# Generate data
set.seed(13)
x <- matrix(rnorm(1000),100,10)
y <- rnorm(100) + 1 + 0.25*x[,1] + 0.5*x[,2] + 0.5*x[,3] 
data <- as.data.frame(cbind(y,x))
# Fit linear regression models with increasing number of variables
r2 <- rep(NA,10)
r2adj <- rep(NA,10)
for (j in 1:10){
data_temp<-data[,1:(j+1)]
r2[j] <- summary(lm(y ~ ., data=data_temp))$r.squared 
r2adj[j] <- summary(lm(y ~ ., data=data_temp))$adj.r.squared 
}
# Adjust margins of plotting region
par(mar=c(5,6,1,1)+.1)
# Plot sequence of R^2 and adjusted R^2. 
plot(2:10, r2[2:10],type="l",lwd=2,lty=3,col="blue",
     xlab="number of predictor variables",ylab=expression(R^2*", "*R[adj]^2),
     ylim=c(min(r2adj[2:10]),max(r2)), cex.lab=1.5, xaxp=c(2,10,8))
# Add points to illustrate values taken by R^2 and adjusted R^2. 
points(2:10,r2adj[2:10],type="l",lwd=1.5,col="darkred")
points(2:10,r2[2:10],col="blue",pch=20, cex=1.5)
points(2:10,r2adj[2:10],col="darkred",pch=20, cex=1.5)
# Add legend
legend("bottomright", c(expression(R^2), expression(R[adj]^2)), 
       col = c("blue","darkred"), lty = c(3, 1), pch=c(20,20),cex=1.3, lwd=2)
# Separate truly informative predictors from non-informative predictors
abline(v=3)
abline(h=r2adj[3],col="gray50",lty=2)
```

This time, as we add predictor variables one at a time to the model we
observe that $R^2$ increases rapidly when the first three informative
predictor variables are added. It continues to increase albeit at a
slower rate as we add the remaining predictor variables one by one.

Note the contrast in behaviour by the adjusted $R^2$, which also rapidly
increases as the first three informative predictor variables are added.
However, as the non-informative predictor variables are added, it stops
increasing and at times even decreases. This is because the adjusted
$R^2$ not only considers the reduction in residual sum of squares as we
move from one model to the next, but also penalises for model
complexity. It is thus more effective as a model selection statistic.

```{r, echo=F, fig.cap="Figure 3: Extra graph about the first experiment."}
# Generate data
set.seed(22)
x <- matrix(rnorm(5000),100,50)
y <- rnorm(100)
data <- as.data.frame(cbind(y,x))
# Fit linear regression models with increasing number of variables
r2 <- rep(NA,50)
r2adj <- rep(NA,50)
for (j in 1:50){
data_temp<-data[,1:(j+1)]
r2[j] <- summary(lm(y ~ ., data=data_temp))$r.squared 
r2adj[j] <- summary(lm(y ~ ., data=data_temp))$adj.r.squared 
}
# Plot the sequence of R^2 statistics
plot(r2,type="l",lwd=1.5,col="blue",lty=3,
     xlab="number of predictor variables",ylab=expression(R^2),
     ylim=c(min(r2adj)-0.1, max(r2)+0.1))
points(r2adj,type="l",lwd=1.5,col="darkred")
points(r2,col="blue",pch=20)
points(r2adj,col="darkred",pch=20)
legend("topleft", c(expression(R^2), expression(R[adj]^2)), 
       col = c("blue","darkred"), lty = c(3, 1), pch=c(20,20),cex=1.3, lwd=2)
```
