---
title: "Practical Report 9"
output:
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    highlight: monochrome
  bookdown::gitbook:
    highlight: monochrome
    code_folding: hide
    global_numbering: yes
  html_document:
    df_print: paged
    code_folding: hide
fontsize: 12pt
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
require(knitr)
pdf_doc <- knitr::is_latex_output()
if (pdf_doc) {opts_chunk$set(echo=FALSE, fig.align="center", out.width="60%", comment="")} else {opts_chunk$set(comment="")}
```

# Effectiveness of blood pressure treatment

We consider the model from Practical 7, that is
$${\mathbb E}[\texttt{BPchange}_i]\quad = \quad \beta_0+\beta_1\texttt{BPinit}_i+\beta_2\texttt{Dose}_i, \qquad \mbox{where } i=1, \ldots, 40.$$

```{r, LoadData1}
bpdata <- read.csv("bpdata.csv")
lm.out1 <- lm(BPchange ~ BPinit + Dose, data=bpdata)
```

## Part a

<!--- Perform a test for the existence of regression --->

The test for the existence of regression compares a normal linear model
to the null model. We can perform the test by using the `anova` function
where the first argument is the null model and the second argument is
the model of interest. The output is as follows.

```{r, GlobalFtest}
# Fit null model
lm.out0 <- lm(BPchange ~ 1, data=bpdata)
# Use the anova function with two arguments
anova(lm.out0, lm.out1)
```

<!--- Identify the entries needed to compile the corresponding ANOVA table-->

Note that the output gives the observed $F$-statistic and $p$-value but
we still need to compute the mean squares to construct the full ANOVA
table for the existence of regression. The mean squares are computed as
the ratio between the sum of squares and the corresponding degrees of
freedom. Thus

-   ${\rm RegrMS}  = \frac{{\rm RegrSS}}{{\rm df(RegrSS)}} = \frac{219.527}{2} = 109.762,$
-   ${\rm ResidMS}  = \frac{{\rm ResidSS}}{{\rm df(ResidSS)}} = \frac{42.248}{37} = 1.142,$

We can now summarise this in the standard ANOVA table for the existence
of regression:

```{r, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- " 
| Source   | d.f.| SS  |  MS | F |  
|:--------|:---:|---:|---:|---:|     
| Regression| 2| 219.527 | 109.763  | 96.128 |  
| Residual | 37 | 42.248 | 1.142  | |  
| Total | 39 | 261.775 |  ||  

"
cat(tabl) 
```

Recall that the test for existence of regression is based on
\begin{align*}
H_0:\:  \beta_1=\beta_2=0 \qquad \text{vs} \qquad 
H_1:\:  \mbox{ at least one of } \beta_1,\beta_2 \mbox{ is non-zero}.
\end{align*} The data has $n=40$ cases, the full model has $p=3$
parameters and the null-hypothesis imposes $q=2$ linear constraints.
Thus, under $H_0$, $F \sim F_{q,n-p} = F_{2, 37}$. We can calculate the
p-value given by $P(F_{2, 37} \geq F_{\rm obs}$) where
$F_{\rm obs}=96.128$ and obtain
`r formatC(pf(96.128, df1=2, df2=37, lower.tail=FALSE), format = "e", digits = 2)`
which is substantially smaller than the significance level of 5%. Thus
we have very strong evidence to reject $H_0$. We conclude that at least
one of the two explanatory variables, that is the initial blood pressure
(`BPinit`) and the dosage of the new drug (`Dose`), are informative
about the change in blood pressure (`BPchange`).

We can also get the test for the existence of regression directly from
the output from the `summary()` function applied to the fitted model.
The last line of the output gives the observed $F$-statistic, degrees of
freedom, and $p$-value.

```{r}
summary(lm.out1)
```

## Part b

The `anova()` function with one argument produces a sequential ANOVA
which shows what happens when, starting from the null model (with only
the intercept), we add the terms of the model sequentially from left to
right. If we use the formula
$${\tt BPchange \quad \sim  \quad BPinit + Dose}$$ when defining the
model, then the `R` output shows the effect of adding `BPinit` then
`Dose`.

```{r, LoadData}
anova(lm.out1)
```

## Part c

To test for the effect of `Dose` controlling for `BPinit` the hypotheses
are: \begin{align*}
    H_0:\:  \beta_2 = 0  \quad \text{versus} \quad
    H_1:\:  \beta_2 \neq 0.
  \end{align*}

As `Dose` was entered second in the model, this test is reported on in
the sequential ANOVA in the row corresponding to `Dose`. As the
$p$-value is less than 0.001, there is very strong evidence to reject
$H_0$. We thus conclude that, controlling for initial blood pressure,
the new drug has an effect on blood pressure.

Since $H_0$ sets a single scalar parameter $\beta_2$ equal to zero, we
can also use a $t$-test which is reported on in the summary output of
the model. In fact, this two-sided $t$-test is equivalent to the partial
$F$-test reported on in the sequential ANOVA. The observed $T$-statistic
is equal to -12.284 with a $p$-value of 1.27e-14. (Note that the square
of the observed $T$-statistic is equal to the observed $F$-statistic and
they both have the same $p$-value.)

As a quick review, recall the way that $p$-values are calculated for the
$t$-test. Under $H_0$, $T\sim t_{37}$ where the degrees of freedom is
determined as $n-p= 40-3 = 37$. The observed $T$-statistic is -12.28 and
therefore the corresponding $p$-value
$P(\left|T_{37}\right| \geq \left| T_{obs} \right|) =$ 1.27e-14 which is
computed in `R` as `2 * pt(abs(-12.284), df=37, lower.tail=FALSE)`.

Note the key steps in calculating the p-value for the $t$-test:

-   Use the absolute value of $T_{\rm obs}$ (with function `abs()`).
-   Use the upper tail of the $t$-distribution (`lower.tail=FALSE`).
-   Double the result to get a 2-sided test.

## Part d

The $F$-tests in the `R`-output in Part b are **sequential** tests for
adding first `BPinit` then `Dose`. The test for the existence of
regression in Part a considers the effect of adding both terms
simultaneously. The information we need is in the `R`-output from Part
b; we just need to do some arithmetic.

The sums of squares for the ANOVA table of the test for the existence of
regression are calculated as follows:

-   The total sum of squares is the sum of all 3 entries in the column
    labelled \`\`Sum Sq": $(261.775 = 47.214 + 172.313 + 42.248).$
-   The regression sum of squares is the sum of the two rows for the
    terms `Dose` and `BPinit`: $(219.527 = 47.214 + 172.313).$
-   The residual sum of squares is from the third row labelled
    \`\`Residuals" (42.248.)

Analogously we can compute the degrees of freedom:

-   The total degrees of freedom is the sum of all 3 entries in the
    column labelled "Df":\
    $(39 = 1 + 1 + 37)$
-   The regression degrees of freedom is the sum of the two rows for the
    terms `Dose` and `BPinit`: $(2 = 1 + 1)$
-   The residual degrees of freedom is from the third row labelled
    \`\`Residuals'': (37).

This now allows us to compute the relevant mean squares and the observed
$F$-statistic for the test for the existence of regression. Recall that
a mean square is calculated by dividing the corresponding sum of squares
by its degrees of freedom. Then the observed $F$-statistic is the ratio
of Regression MS to Residual MS. Thus we have

-   ${\rm RegrMS}  = \frac{{\rm RegrSS}}{{\rm df(RegrSS)}} = \frac{219.527}{2} = 109.762,$
-   ${\rm ResidMS}  = \frac{{\rm ResidSS}}{{\rm df(ResidSS)}} = \frac{42.248}{37} = 1.142,$
-   ${ F}_{\rm obs} = \frac{\rm RegrMS}{\rm ResidMS} = \frac{109.762}{1.142} = 96.128.$

We can now summarise this in the standard ANOVA table for the existence
of regression:

```{r, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- " 
| Source   | d.f.| SS  |  MS | F |  
|:--------|:---:|---:|---:|---:|     
| Regression| 2| 219.527 | 109.763  | 96.128 |  
| Residual | 37 | 42.248 | 1.142  | |  
| Total | 39 | 261.775 |  ||  

"
cat(tabl) 
```

# Model selection statistics

In Experiment 1 we generate data for 50 predictor variables
$X_1, \ldots, X_{50}$ all of which are non-informative about the
response $Y$ as the latter is sampled independently from the predictor
variables. We then consider a sequence of models, starting with a simple
linear regression of $Y$ on $X_1$ and adding the other predictor
variables one at a time. For each model the coefficient of determination
$R^2$ is computed. The plot below shows the resulting sequence of $R^2$
values.

```{r, echo=F, fig.cap="A graphical illustration of the coefficient of determination as the number of non-informative predictor variables increases in the model.", fig.alt="A graphical illustration of R squared as the number of non-informative predictor variables increases in the model. We see that R squared has a tendency to increase as the number of predictor variables increases."}
# Generate data
set.seed(22)
x <- matrix(rnorm(5000),100,50)
y <- rnorm(100)
data <- as.data.frame(cbind(y,x))
# 
r2 <- rep(NA,50)
for (j in 1:50){
  data_temp<-data[ ,1:(j+1)]
  r2[j] <- summary(lm(y ~ ., data=data_temp))$r.squared 
}
# Plot the sequence of R^2 statistics
plot(r2,type="l",lwd=1.5,col="blue",
     xlab="number of predictor variables",ylab=expression(R^2))
points(r2,col="blue",pch=20)
```

We observe that despite the predictor variables all being
non-informative about the response, the coefficient of determination
$R^2$ continues to increase as we add more predictor variables to the
model.

In Experiment 2 we generate data for 10 predictor variables. The
response is sampled from the model
$$Y_i = 1 + 0.25x_{i,1}  + 0.5x_{i,2} + 0.5 x_{i,3}  + \epsilon_i, \qquad \mbox{where }  \epsilon_i \sim {\cal N}(0,1) \;  {\rm iid},  \qquad i=1, \ldots, 10.$$
Thus the first three predictor variables are informative about the
response, while the other predictor variables are not.

```{r, echo=F, fig.cap="A graphical illustration of the difference of behaviour between the coefficient of determination and the adjusted coefficient of determination as the number of predictor variables increases in the model.", fig.alt="Another graphical illustration of the difference of behaviour between R squared and adjusted R squared as the number of predictor variables increases in the model. We see that both R squared and adjusted R squared increase as informative predictor variables are added to the model, but then R squared continues to increase as non-informative predictor variables are added while R squared adjusted tends to decrease."}
# Generate data
set.seed(13)
x <- matrix(rnorm(1000),100,10)
y <- rnorm(100) + 1 + 0.25*x[,1] + 0.5*x[,2] + 0.5*x[,3] 
data <- as.data.frame(cbind(y,x))
#
r2 <- rep(NA,10)
r2adj <- rep(NA,10)
for (j in 1:10){
data_temp<-data[,1:(j+1)]
r2[j] <- summary(lm(y ~ ., data=data_temp))$r.squared 
r2adj[j] <- summary(lm(y ~ ., data=data_temp))$adj.r.squared 
}
# Adjust margins of plotting region
par(mar=c(5,6,1,1)+.1)
# Plot sequence of R^2 and adjusted R^2. 
plot(2:10, r2[2:10],type="l",lwd=2,lty=3,col="blue",
     xlab="number of predictor variables",ylab=expression(R^2*", "*R[adj]^2),
     ylim=c(min(r2adj[2:10]),max(r2)), cex.lab=1.5, xaxp=c(2,10,8))
# Add points to illustrate values taken by R^2 and adjusted R^2. 
points(2:10,r2adj[2:10],type="l",lwd=1.5,col="darkred")
points(2:10,r2[2:10],col="blue",pch=20, cex=1.5)
points(2:10,r2adj[2:10],col="darkred",pch=20, cex=1.5)
# Add legend
legend("bottomright", c(expression(R^2), expression(R[adj]^2)), 
       col = c("blue","darkred"), lty = c(3, 1), pch=c(20,20),cex=1.3, lwd=2)
# Separate truly informative predictors from non-informative predictors
abline(v=3)
abline(h=r2adj[3],col="gray50",lty=2)
```

This time, as we add predictor variables one at a time to the model we
observe that $R^2$ increases rapidly when the first three informative
predictor variables are added. It continues to increase albeit at a
slower rate as we add the remaining predictor variables one by one.

Note the contrast in behaviour by the adjusted $R^2$, which also rapidly
increases as the first three informative predictor variables are added.
However, as the non-informative predictor variables are added, it stops
increasing and at times even decreases. This is because the adjusted
$R^2$ not only considers the reduction in residual sum of squares as we
move from one model to the next, but also penalises for model
complexity. It is thus more effective as a model selection statistic.
