---
title: "Practical Report 9"
output:
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    highlight: monochrome
  bookdown::gitbook:
    highlight: monochrome
    code_folding: hide
    global_numbering: yes
  html_document:
    df_print: paged
    code_folding: hide
fontsize: 12pt
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
require(knitr)
pdf_doc <- knitr::is_latex_output()
if (pdf_doc) {opts_chunk$set(echo=FALSE, fig.align="center", out.width="60%", comment="")} else {opts_chunk$set(comment="")}
```

# Effectiveness of blood pressure treatment

We begin with the model from Practical 7, that is $$
{\mathbb E}[\texttt{BPchange}_i]\quad = \quad \beta_0+\beta_1\texttt{BPinit}_i+\beta_2\texttt{Dose}_i, \qquad \mbox{where } i=1, \ldots, 40.
$$

```{r, LoadData}
bpdata <- read.csv("../data/bpdata.csv")
lm.out1 <- lm(BPchange ~ BPinit + Dose, data=bpdata)
```

## Part a

We perform the test by `anova` between null model and `lm.out1`.

```{r, GlobalFtest}
# Fit null model
lm.out0 <- lm(BPchange ~ 1, data=bpdata)
# Use the anova function with two arguments
anova(lm.out0, lm.out1)
```

| Source | d.f. | SS | MS | F |
|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
| Regression | 2 | $\sum_{i=1}^{n} (\hat{y}_i - \bar{y})^2$ | $\frac{\text{RegrSS}}{p-1}$ | 96.128 |
| Residual | 37 | $\sum_{i=1}^{n} (y_i - \hat{y}_i)^2$ | $\frac{\text{ResidSS}}{n-p}$ |  |
| Total | 39 | $\sum_{i=1}^{n} (y_i - \bar{y})^2$ |  |  |

The output gives the observed $F$-statistic and $p$-value but we still need to compute the mean squares and sum of squares to construct the full ANOVA table for the existence of regression.

From the table, we know directly that residual sum of squares (ResidSS) is
$$
\sum_{i=1}^n (y_i - \hat{y}_i)^2 = 42.248,
$$
and the regression sum of squares (RegrSS) is
$$
\begin{align*}
\sum_{i=1}^n (\hat{y}_i - \bar{y})^2 &= \sum_{i=1}^n (y_i - \bar{y})^2 - \sum_{i=1}^n (y_i - \hat{y}_i)^2\\
&= \text{ResidSS of Model 1} - \text{ResidSS of Model 2}\\
&= 261.775 - 42.248 = 219.527
\end{align*}
$$

For the mean squares, we have
-  ${\rm RegrMS}  = \frac{{\rm RegrSS}}{{\rm df(RegrSS)}} = \frac{219.527}{2} = 109.762,$
-  ${\rm ResidMS}  = \frac{{\rm ResidSS}}{{\rm df(ResidSS)}} = \frac{42.248}{37} = 1.142,$

## Part b

<!--- Perform a sequential ANOVA --->

```{r, SequentialAnova}

```

## Part c

<!--- Controlling for initial blood pressure, does the new drug have an effect on blood pressure? --->

## Part d

<!--- How can the results of the sequential ANOVA be used to re-produce the results for the ANOVA for a test of existence in a? --->

# Model selection statistics

<!--- Experiment 1 --->

```{r,  fig.cap=""}
# Generate data
set.seed(22)
x <- matrix(rnorm(5000),100,50)
y <- rnorm(100)
data <- as.data.frame(cbind(y,x))
# 
r2 <- rep(NA,50)
for (j in 1:50){
  data_temp<-data[ ,1:(j+1)]
  r2[j] <- summary(lm(y ~ ., data=data_temp))$r.squared 
}
# Plot the sequence of R^2 statistics
plot(r2,type="l",lwd=1.5,col="blue",
     xlab="number of predictor variables",ylab=expression(R^2))
points(r2,col="blue",pch=20)
```

<!--- Experiment 2 -->

```{r, fig.cap=""}
# Generate data
set.seed(13)
x <- matrix(rnorm(1000),100,10)
y <- rnorm(100) + 1 + 0.25*x[,1] + 0.5*x[,2] + 0.5*x[,3] 
data <- as.data.frame(cbind(y,x))
#
r2 <- rep(NA,10)
r2adj <- rep(NA,10)
for (j in 1:10){
data_temp<-data[,1:(j+1)]
r2[j] <- summary(lm(y ~ ., data=data_temp))$r.squared 
r2adj[j] <- summary(lm(y ~ ., data=data_temp))$adj.r.squared 
}
# Adjust margins of plotting region
par(mar=c(5,6,1,1)+.1)
# Plot sequence of R^2 and adjusted R^2. 
plot(2:10, r2[2:10],type="l",lwd=2,lty=3,col="blue",
     xlab="number of predictor variables",ylab=expression(R^2*", "*R[adj]^2),
     ylim=c(min(r2adj[2:10]),max(r2)), cex.lab=1.5, xaxp=c(2,10,8))
# Add points to illustrate values taken by R^2 and adjusted R^2. 
points(2:10,r2adj[2:10],type="l",lwd=1.5,col="darkred")
points(2:10,r2[2:10],col="blue",pch=20, cex=1.5)
points(2:10,r2adj[2:10],col="darkred",pch=20, cex=1.5)
# Add legend
legend("bottomright", c(expression(R^2), expression(R[adj]^2)), 
       col = c("blue","darkred"), lty = c(3, 1), pch=c(20,20),cex=1.3, lwd=2)
# Separate truly informative predictors from non-informative predictors
abline(v=3)
abline(h=r2adj[3],col="gray50",lty=2)
```
