---
title: "Practical Report 7"
output:
   pdf_document: 
     keep_tex: true
     latex_engine: xelatex
     fig_caption: yes
     highlight: monochrome
   html_document:
    df_print: paged
    code_folding: hide
fontsize: 12pt
---

```{r setup, include = FALSE}
require(knitr)
pdf_doc <- knitr::is_latex_output()
if (pdf_doc) {opts_chunk$set(echo=FALSE, fig.align="center", out.width="60%")} else {opts_chunk$set(echo=TRUE)}
```

# Effectiveness of blood pressure treatment {-}

a. Before fitting linear models we first explore the data graphically to assess whether a linear model is appropriate.
```{r, LoadData, echo=FALSE}
bpdata <- read.csv("bpdata.csv")
```

```{r, Plot1, echo=FALSE, fig.cap="A scatterplot of change in blood pressure against the initial blood pressure."}

```

```{r, Plot2, echo=FALSE, fig.cap ="A scatterplot of change in blood pressure (BPchange) against the dose of the new drug (Dose). The observations are slightly jittered."}

```

<!--- Comment on the plots --->


b. We first fit the model  $${\mathbb E}[\texttt{BPchange}_i]\quad = \quad \beta_0+\beta_1\texttt{BPinit}_i+\beta_2\texttt{Dose}_i$$
where $i =1, \ldots, 40.$

```{r, 1b,  echo=FALSE}
lm.out <- 
```  

c. Next we convert `Dose` to a factor which allows us to model a non-linear dose-response relationship before fitting a new model.

```{r, 1c,  echo=FALSE}
lm.factor <- 
summary(lm.factor)
```
   
    1. The formula for the new model is:

    2. <!--- Interpretation of parameter --->

    3. <!--- Comparison of estimates for factor Dose and corresponding estimate from the linear model --->
  
> A side-by-side comparison of the estimates is contained in the table below:

  
```{r, echo=TRUE}
# Expected blood pressure change for each dose in the linear model
BPmean <- mean(bpdata$BPinit)
DoseF <- c(10,20,30,40)
clin <- predict(lm.out, newdata=list(Dose=DoseF, BPinit=rep(BPmean, 4)))
    
# Expected blood pressure change for each dose in the factor model
cfac <- coef(lm.factor)[1:4]

# Create the required table
require("knitr")
tabl <- cbind("Dose"=DoseF,"Factor Model"=cfac, "Linear Model"=clin)
rownames(tabl)<-NULL
kable(round(tabl, 1))
```  

d. Next we consider the collaborator's enquiry.

    1. <!--- CI for the effect of a 50 50$\mu$g dose --->
    
    2. <!--- What does the factor model tell us about the effect of a 50$\mu$g dose?---> 
    
    3. <!--- Consider a  100$\mu$g dose ---> 




# Polynomial regression {-}

a. We plot the data and then add the fitted curve from a cubic regression model.

    ```{R, cubic, echo=TRUE, fig.cap="Scatterplot of the polynomial data. Added to the plot is a fitted curbe from a polynomial regression model."}
#
poldata <- read.csv("poldata.csv")
m3 <- lm(y ~ x + I(x^2) + I(x^3), data=poldata)
# 
ax <- seq(from=2, to=6, length=101)
# 
p1 <- predict(m3, newdata=list(x=ax))
# 
plot(y ~ x, data=poldata, ylim=range(p1), pch=16)
lines(ax, p1)
    ```

<!---  Comment on the plot --->


b. <!--- Confidence and prediction intervals --->

    ```{R, echo=TRUE, fig.cap="" }
# 
p2 <- predict(m3, newdata=list(x=ax), interval="confidence")
p3 <- predict(m3, newdata=list(x=ax), interval="prediction")
# 
plot(y ~ x, data=poldata, ylim=range(p3), pch=16)
# 
matlines(ax, p2, col=c("black", "blue", "blue"), lty=c(1,3,3))
matlines(ax, p3, col=c("black", "red", "red"), lty=c(1,2,2))
#
legend("topright", legend=c("fitted mean function", 
              "95% confidence interval", "95% prediction interval"), 
       col=c("black", "blue", "red"), lty=c(1,3,2), cex=0.9)
    ```


c. <!--- Prediction interval for extended x-axis --->

```{r, echo=F, fig.cap=""}
#
ax <- seq(from=0, to=6, length=101)
#  
p <- predict(m3, newdata=list(x=ax), interval="prediction")
#
plot(y ~ x, xlim=c(0, 6), data=poldata, ylim=range(p))
#
matlines(ax, p, col="black", lty=c(1,2,2))
```


d. <!--- Predicting response for x=0? --->


