---
title: "Practical Report 7"
output:
   pdf_document: 
     keep_tex: true
     latex_engine: xelatex
     fig_caption: yes
     highlight: monochrome
   html_document:
    df_print: paged
    code_folding: hide
fontsize: 12pt
---

```{r setup, include = FALSE}
require(knitr)
pdf_doc <- knitr::is_latex_output()
if (pdf_doc) {opts_chunk$set(echo=FALSE, fig.align="center", out.width="60%")} else {opts_chunk$set(echo=TRUE)}
```

# Effectiveness of blood pressure treatment {.unnumbered}

## A

Before fitting linear models we first explore the data graphically to assess whether a linear model is appropriate.

```{r, LoadData, echo=FALSE}
bpdata <- read.csv("../data/bpdata.csv")
```

```{r, Plot1, echo=FALSE, fig.cap="A scatterplot of change in blood pressure against the initial blood pressure."}
plot(BPchange ~ BPinit, data=bpdata, pch=16, ylim = c(-20,0),
     xlab="Initial blood pressure (mmHg)", 
     ylab="Change in blood pressure (mmHg)")
```

For all patients the blood pressure was lower in the second measurement as all values of the change in blood pressure are negative. The scatterplot shows that with increasing initial blood pressure the reduction in blood pressure becomes larger. The relationship between initial blood pressure and the change of blood pressure appears linear. The observed variation is smaller for initial blood pressures less than 120 mmHg, but this could be due to the small number of observations in this range.

```{r, Plot2, echo=FALSE, fig.cap ="A scatterplot of change in blood pressure against the dose of the new drug."}
plot(BPchange ~ Dose, data=bpdata, pch=16, 
     xlab=expression(paste("Dose (", mu,"g)")), 
     ylab="Change in blood pressure (mmHg)")
```

```{r, Plot2_2, echo=FALSE, fig.cap ="A scatterplot of change in blood pressure (BPchange) against the dose of the new drug (Dose). The observations are slightly jittered."}
plot(jitter(BPchange, amount=0.2) ~ Dose, data=bpdata, 
     xlab=expression(paste("Dose (", mu,"g)")), 
     ylab="Change in blood pressure (mmHg)")
```

Jittering the response values which adds a little noise to the datapoints. We observe that the reduction in blood pressure tends to become larger as the dose increases and it appears reasonable to assume a linear relationship between blood pressure and dose. We observe that the spread of the response appears larger for a dose of 30$\mu$g than for the other dosages.

## B

We first fit the model $${\mathbb E}[\texttt{BPchange}_i]\quad = \quad \beta_0+\beta_1\texttt{BPinit}_i+\beta_2\texttt{Dose}_i$$ where $i =1, \ldots, 40.$

```{r, 1b,  echo=FALSE}
lm.out <- lm(BPchange ~ BPinit + Dose, data=bpdata)
summary(lm.out)
```

## C

Next we convert `Dose` to a factor which allows us to model a non-linear dose-response relationship before fitting a new model.

```{r, 1c, echo=FALSE}
lm.factor <- lm(BPchange ~  0 + factor(Dose) + I(BPinit - mean(BPinit)), 
                data=bpdata)
summary(lm.factor)
```

1.  The mathematical expression is $$
    {\mathbb E}[\texttt{BPchange}_i]\quad  = \quad \alpha_{\text{Dose}_i}+\beta_1 (\texttt{BPinit}_i - \overline{\texttt{BPinit}}).
    $$
2.  The parameter $\alpha_k$ is the expected change in blood pressure for a person with average initial blood pressure (second term is zero) given dose $k$.

A side-by-side comparison of the estimates is contained in the table below:

```{r}
# Expected blood pressure change for each dose in the linear model
BPmean <- mean(bpdata$BPinit)
DoseF <- c(10,20,30,40)
clin <- predict(lm.out, newdata=list(Dose=DoseF, BPinit=rep(BPmean, 4)))

# Expected blood pressure change for each dose in the factor model
cfac <- coef(lm.factor)[1:4]

# Create the required table
require("knitr")
tabl <- cbind("Dose"=DoseF,"Factor Model"=cfac, "Linear Model"=clin)
rownames(tabl)<-NULL
kable(round(tabl, 1))
```

## D

Next we consider the collaborator's enquiry.

1.  The coefficient of `Dose` is the expected change in blood pressure for a 1$\mu$g dose. To estimate the effect of a 50$\mu$g dose we need to multiply the coefficient by 50. To avoid spurious precision, we round to one decimal place. Thus, controlling for initial blood pressure, the effect of a 50$\mu$g dose is an expected reduction in blood pressure by

```{r}
-as.numeric(round(50 * coef(lm.out)["Dose"], 1))
```

2.  The factor model cannot tell us anything about the effect of a 50$\mu$g dose. The effect of factor variables cannot be extrapolated beyond the levels observed when the model is fitted.

3.  A 100$\mu$g dose is well beyond the range of our observed data. There is no reason to be confident that the dose-response effect will be linear this far beyond our observation.

# Polynomial regression

## A

We plot the data and then add the fitted curve from a cubic regression model. We observe a non-linear and non-monotonic relationship between the explanatory variable and the response which we model using a cubic regression model.

```{R, cubic, echo=TRUE, fig.cap="Scatterplot of the polynomial data. Added to the plot is a fitted curbe from a polynomial regression model."}
# Read data
poldata <- read.csv("../data/poldata.csv")
# Fit polynomial model
m3 <- lm(y ~ x + I(x^2) + I(x^3), data=poldata) #Cubic model
# Plot of the data
plot(y ~ x, data=poldata, pch=16)
# Vector of x values at which to compute fitted y value
ax <- seq(from=2, to=6, length=101)
# Compute fitted values 
p <- predict(m3, newdata=list(x=ax))
# Plot smooth line through fitted values
lines(ax, p)
```

## B

Next, for the cubic model we illustrate a 95% confidence interval for the mean function and a 95% prediction interval for every value of $x$. We observe that the confidence interval is narrower than the corresponding prediction interval.

-   Confidence interval estimates the uncertainty around the mean value of the dependent variable for a given value of the independent variable.
-   Prediction interval predicts the range where one specific future data point will fall.

```{R, echo=TRUE, fig.cap="" }
# Create a sequence of x-values
ax <- seq(from=2, to=6, length=101)
# compute the confidence interval and the prediction interval for each x-value
p2 <- predict(m3, newdata=list(x=ax), interval="confidence")
p3 <- predict(m3, newdata=list(x=ax), interval="prediction")
# plot the data such that all prediction intervals fit on the plot
plot(y ~ x, data=poldata, ylim=range(p3), pch=16)
# plot smooth lines through the prediction interval endpoints
matlines(ax, p2, col=c("black", "blue", "blue"), lty=c(1,3,3))
matlines(ax, p3, col=c("black", "red", "red"), lty=c(1,2,2))
legend("topright", 
       legend=c("fitted mean function", 
            "95% confidence interval", 
            "95% prediction interval"), 
       col=c("black", "blue", "red"), lty=c(1,3,2), cex=0.9)
```

## C

We observe that the prediction interval becomes wider as we move away from the centre of the data.

```{r, echo=F, fig.cap=""}
# Create a sequence of x-values
ax <- seq(from=0, to=6, length=101)
# compute the prediction interval for each x-value
p <- predict(m3, newdata=list(x=ax), interval="prediction")
# plot the data such that all prediction intervals fit on the plot
plot(y ~ x, xlim=c(0, 6), data=poldata, ylim=range(p), pch=16)
# plot smooth lines through the prediction interval endpoints
matlines(ax, p, col="black", lty=c(1,2,2))
```

## D

If the model is true, that is it represents the true data-generating mechanism, then the prediction is valid. However, we do not know if it is true. One reason to be skeptical is that the cubic model shows a maximum around $x=2$ and then a decrease for smaller values of $x$, despite no evidence from the data that there is a turning point in the predictor-response relationship. Note that this behaviour is built into the cubic model.
