---
title: "Lab 5 solution"
author: "Haochu"
format:
  revealjs:
    code-fold: false
    scrollable: true
    fontsize: 20pt
---

## Question 1

Given two models:

-   `Model 1: log(price) ~ log(weight)`
-   `Model 2: log(price) ~ log(weight) + cut`

## Question 1

@fig-1 shows two scatterplots of residuals against fitted values. The left plot that the residuals for fair-cut diamonds tend to be negative. This systematic structure is resolved in the right plot where the residuals for each cut category scatter more evenly above and below the smoother. Hence Model 2 is more appropriate here.

```{r Q1}
#| label: fig-1
#| fig-cap: "Residuals versus fitted values plots. Left: Model 1. Right: Model 2."

library(RColorBrewer)

dia5 <- read.csv("../data/dia5.csv")
lm1.out <- lm(log(price) ~ log(weight), data=dia5)
lm2.out <- lm(log(price) ~ log(weight) + cut, data=dia5)
ind <- ifelse(dia5$cut=="Fair", 1, ifelse(dia5$cut == "Very Good", 2, 3))
col <-  brewer.pal(3, "Dark2")

par(mfrow = c(1, 2))

plot(lm1.out, 1, col=col[ind], pch=ind)
legend("topright", legend=c("Fair", "Very Good", "Ideal"), pch=c(1,2,3), col=col, title="diamond cut", cex=0.75)
plot(lm2.out, 1, col=col[ind], pch=ind)
legend("topright", legend=c("Fair", "Very Good", "Ideal"), pch=c(1,2,3), col=col, title="diamond cut", cex=0.75)

par(mfrow = c(1, 1))
```

## Question 2

-   `Model 2: log(price) ~ log(weight) + cut`
-   `Model 3: log(price) ~ log(weight) + cut + log(weight):cut`

## Question 2

@fig-2 below shows scatterplots of standardised residuals against log-weight for fair cut diamonds. We observe in the left plot that the residuals for small diamonds tend to be positive while for larger diamonds they tend to be negative. This is also shown by the negative slope of the line of best fit and thus indicates that there is systematic structure within the data that is not explained by Model 2. In Model 3, the residuals are relatively evenly distributed above and below the line, thus showing no systematic pattern. The plots support the inclusion of an interaction between cut and log-weight in the model. Thus Model 3 is preferable to Model 2, least for fair cut diamonds.

```{r Q2}
#| label: fig-2
#| fig-cap: "Scatterplots standardised residuals against log-weight for fair cut diamonds for Model 2 (left) and Model 3 (right)."

lm3.out <- lm(log(price) ~ log(weight) + cut + cut:log(weight), data=dia5)
res2 <- rstandard(lm2.out)
res3 <- rstandard(lm3.out)
ind <- dia5$cut=="Fair"

par(mfrow = c(1, 2))

plot(res2[ind] ~ log(dia5$weight[ind]), xlab="log-weight (in log-carat)", ylab="Standardised residuals")
abline(lm(res2[ind] ~ log(dia5$weight[ind])), col="red")
plot(res3[ind] ~ log(dia5$weight[ind]), xlab="log-weight (in log-carat)", ylab="Standardised residuals")
abline(lm(res3[ind] ~ log(dia5$weight[ind])), col="red")

par(mfrow = c(1, 1))
```

## Question 3

Our analysis has shown that, in addition to the weight of a diamond, its cut is informative about its price. If the weight of diamonds are the same, better cut quality diamonds tend to have higher prices. And for a given percentage increase in the weight of a diamond, the resulting percentage increase in the predicted price depends on the cut of the diamond.
