---
title: "Practical Report 8"
output:
   pdf_document: 
     keep_tex: true
     fig_caption: yes
     highlight: monochrome
   html_document:
    df_print: paged
    code_folding: hide
fontsize: 12pt
---

```{r setup, include = FALSE}
require(knitr)
pdf_doc <- knitr::is_latex_output()
if (pdf_doc) {opts_chunk$set(echo=FALSE, fig.align="center", out.width="70%")} else {opts_chunk$set(echo=TRUE)}
```




# Polynomial regression 



a. In Practical 7 we observed a non-linear and non-monotonic relationship between the explanatory variable and the response. 
We fit four polynomial regression models of increasing complexity to the data. 

```{r, fig.cap=""}
#
poldata <- read.csv("poldata.csv")
plot(y ~ x, data=poldata, pch=16)
# 
m1 <- lm(y ~ x,                            data=poldata) #Linear model
m2 <- lm(y ~ x + I(x^2),                   data=poldata) #Quadratic model
m3 <- lm(y ~ x + I(x^2) + I(x^3),          data=poldata) #Cubic model
m4 <- lm(y ~ x + I(x^2) + I(x^3) + I(x^4), data=poldata) #Quartic model
# 
ax <- seq(from=2, to=6, length=101)
# We save the models in a list and then use a for-loop 
# over the items in the list to save on writing
mlist <- list(m1, m2, m3, m4)
for (i in seq_along(mlist)) {
  #
  p <- predict(mlist[[i]], newdata=list(x=ax))
  #
  lines(ax, p, col=5-i, lwd=2, lty=i)
}
#
legend("topright", lwd=2, col=4:1, legend=1:4, lty=1:4, title="degree")
```

<!---  Comment on the plot --->

b. An unbiased estimate for the error variance can be computed as ...

c.  <!--- Hypothesis test to decide whether a quartic term is needed in the model--->

<!--- Note how to report on a hypothesis test
- state the null and alternative hypothesis;
- give an algebraic expression of the test statistic;
- give the distribution of the test statistic under the null hypothesis;
- give the observed value of the test statistic;
- report the $p$-value;
- state the outcome of the test;
- draw an appropriate  conclusion.
--->


d. <!--- Answer to question d. --->

e. <!--- Hypothesis test: quadratic versus cubic model --->


# Multicollinearity

In the following we illustrate the issues that arise when we have multi-collinear predictors. We start by simulating realisations of 3 explanatory variables $X_1$, $X_2$ and $X_3$, such that $X_1$ and $X_3$ have a strong linear relationship. We then sample realisations of the response variable from the model
\[Y_i \quad = \quad 1 + x_{i1} + x_{i2} + x{i_3} + \epsilon_i,\]
where $i=1, \ldots, 100$ and $\epsilon_1, \ldots, \epsilon_{100}$ are iid N(0, $0.5^2$) distributed.

```{r}
# Producing realisations from the explanatory variables
set.seed(8)
x1 <- runif(100)
x2 <- runif(100)
x3 <- 2 * x1 + rnorm(100,0,0.02)
# Sampling response observations
y  <- 1 + x1 + x2 + x3 + rnorm(100,0,0.5)
```

We can illustrate the relationship between the explanatory variables using a `pairs` plot. 

```{r,out.width="65%",fig.align='center', fig.cap="A scatterplot matrix of the three explanatory variables x1, x2 and x3 from the artificial data set."}
pairs(x1 ~ x2 + x3)
```

<!--- Comment on the scatterplot matrix ---> 


First we consider a linear model of $Y$ on all three explanatory variables, that is $X_1$, $X_2$ and $X_3$. 

```{r}
m2 <- lm(y ~ x1 + x2 + x3)
summary(m2)
```

<!--- Comment on the results ---> 


Next consider a linear model of $Y$ on $X_2$ and $X_3$. 

```{r}
m1 <- lm(y ~ x1 + x2)
summary(m1)
```

<!--- Comment on the results---> 


We can quantify multi-collinearity using variance inflation factors. The $j$th variance inflation factor ${\rm VIF}_j$ is the multiplicative factor by which the variance of the estimator $\hat{\beta_j}$ is inflated due to (approximate) collinearity.

The variance inflation factors  be computed using `vif` from car package. As a rule of thumb we regard a ${\rm VIF}_j> 10$ as a concern.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(car)
round(vif(m2), 2)
```
