---
title: "Practical Report 8"
output:
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    latex_engine: xelatex
    highlight: monochrome
  html_document:
    df_print: paged
    code_folding: hide
fontsize: 12pt
---

```{r setup, include = FALSE}
require(knitr)
pdf_doc <- knitr::is_latex_output()
if (pdf_doc) {opts_chunk$set(echo=FALSE, fig.align="center", out.width="60%", comment="")} else {opts_chunk$set(comment="")}
```


# Polynomial regression {-}


## Part a

In Practical 7 we observed a non-linear and non-monotonic relationship between the explanatory variable and the response. 
We fit four polynomial regression models of increasing complexity to the data. 
The fitted curves from the four models are shown in the Figure 1. Since we are performing repetitive calculations, we assign the fitted models to a list and then iterate over the elements of the list in a `for` loop. This is implemented in the code chunk below. 


```{r, echo=T, fig.cap="A scatterplot of the data. Also shown are the fitted curves of the four polynomial regression models, ranging from degree 1 to degree 4.", fig.alt="A scatterplot of the data showing a curvilinear relationship between the predictor x and response y. The fitted lines/curves for the polynomial models of degree 1 to degree 4 are added to the plot. The fit of the simple linear regression line is inadequate. The The fitted cubic and quartic model are virtually indistinguishable in the plot, but appear to provide a better fit than the quadratic model.", out.width="70%", fig.align='center'} 
# Load and plot data 
poldata <- read.csv("poldata.csv")
plot(y ~ x, data=poldata, pch=16)
# Fit the four polynomial models
m1 <- lm(y ~ x,                            data=poldata) #Linear model
m2 <- lm(y ~ x + I(x^2),                   data=poldata) #Quadratic model
m3 <- lm(y ~ x + I(x^2) + I(x^3),          data=poldata) #Cubic model
m4 <- lm(y ~ x + I(x^2) + I(x^3) + I(x^4), data=poldata) #Quartic model

# Vector of x values at which to compute  predicted y value
ax <- seq(from=2, to=6, length=101)
# Assign fitted models to a list
mlist <- list(m1, m2, m3, m4)
# Loop over list of models
for (i in seq_along(mlist)) {
    # Compute fitted values
    p <- predict(mlist[[i]], newdata=list(x=ax))
    # Plot smooth line through fitted values
    lines(ax, p, col=5-i, lwd=2, lty=i)
}
# Add a legend
legend("topright", lwd=2, col=4:1, legend=1:4, lty=1:4, title="degree")
```

\newpage

Notice in Figure 1 that the fitted cubic and quartic model are virtually indistinguishable in the plot.


## Part b
We can either plot the whole model summary as below or extract the residual standard error $\sqrt{s^2}$ using the command `summary(m4)$sigma`.  

```{r}
summary(m4)
```

`R` reports as the residual standard error $\sqrt{s^2}=$ `r round(summary(m4)$sigma, 3)` and so the unbiased estimate for the error variance is $s^2 =$ `r round(summary(m4)$sigma^2, 4)`.

## Part c
Let the formula for the
quartic model be represented algebraically as
\[
Y \quad = \quad  \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \beta_4 x^4 + \epsilon.
\]
To test 
$H_0: \beta_4 = 0$ versus $H_1: \beta_4 \neq 0$ we use a $t$-test. 
The number of units of observation in the dataset is $n=14$ and the quartic model has $p=5$ parameters. Hence, under $H_0$, the $T$-statistic, defined as $\frac{\widehat{\beta_4}}{{\rm std.error}(\widehat{\beta_4})}$ has a $t$-distribution on $n-p=14-5 = 9$ degrees of freedom. As is listed in the 4th row of the table of coefficients, the observed value of the $T$-statistic is $T_{\rm obs} = -0.079$ which has a $p$-value of $0.939$. Thus there is insufficient evidence at the 5\% level to reject the null hypothesis that $\beta_4=0$. We conclude that the quartic model is not a significant improvement over the cubic model. Hence we do not include a quartic term in the model and instead retain the cubic model.

**Note how we reported on the hypothesis test.** We

- stated the null and alternative hypothesis;
- defined the test statistic and gave its distribution under the null hypothesis;
- gave the observed value of the test statistic;
- reported the $p$-value;
- stated the outcome of the test;
- drew the corresponding conclusion.

## Part d
It is not possible to read from the model summary for the quartic model whether a cubic polynomial is needed.  The $T$-statistics used for the model summary of the polynomial regression of degree 4 are not independent. The $p$-value reported for the cubic term may change after removing the quartic term from the model. 

## Part e
Let the formula for the
cubic model be represented algebraically as
\[
Y = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \epsilon.
\]
We would like to test 
$H_0: \beta_3 = 0$ versus $H_1: \beta_3 \neq 0$. 
The number of units of observation in the dataset is $n=14$ and the cubic model has $p=4$ parameters. Hence, under $H_0$, the $T$-statistic, defined as $\frac{\widehat{\beta_3}}{{\rm std.error}(\widehat{\beta_3})}$  has a $t$-distribution with $n-p=14-4 = 10$ degrees of freedom. The observed value of the $T$-statistic is 
$$T_{\rm obs}  \quad = \quad \frac{0.059}{0.009} \quad = \quad  6.56.$$ The associated $p$-value is given by
\[ p \quad = \quad  2 \Big[1- F_{t_{10}}(6.56)\Big] \quad < \quad  6.4 \times 10^{-5}.\]
(We can compute $F_{t_{10}}(6.56)$ using the command `pt(q=6.56, df=10)`.)
The $p$-value shows that there is very strong evidence against the null hypothesis that $\beta_3=0$. We conclude that the cubic model is a significant improvement over the quadratic model at the 5\% level and so a cubic term is needed in the model. 
The result above matches the ones reported in the model summary (modulo some rounding errors), see below. 

```{r}
if (!pdf_doc){summary(m3)}
```

```{ =latex}
\begin{verbatim}
   Call:
   lm(formula = y ~ x + I(x^2) + I(x^3), data = poldata)

   Coefficients:
                Estimate Std. Error t value Pr(>|t|)    
   (Intercept) -0.430626   0.503431  -0.855 0.412365    
   x            1.670739   0.416908   4.007 0.002488 ** 
   I(x^2)      -0.608420   0.109077  -5.578 0.000235 ***
   I(x^3)       0.058812   0.009062   6.490 6.99e-05 ***
   ---
   Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
   \end{verbatim}
```

# Multicollinearity

In the following we illustrate the issues that arise when we have multi-collinear predictors. We start by simulating realisations of 3 explanatory variables $X_1$, $X_2$ and $X_3$, such that $X_1$ and $X_3$ have a strong linear relationship. We then sample realisations of the response variable from the model
\[Y_i \quad = \quad 1 + x_{i1} + x_{i2} + x{i_3} + \epsilon_i,\]
where $i=1, \ldots, 100$ and $\epsilon_1, \ldots, \epsilon_{100}$ are iid N(0, $0.5^2$) distributed.

```{r}
# Producing realisations from the explanatory variables
set.seed(8)
x1 <- runif(100)
x2 <- runif(100)
x3 <- 2 * x1 + rnorm(100,0,0.02)
# Sampling response observations
y  <- 1 + x1 + x2 + x3 + rnorm(100,0,0.5)
```

We can illustrate the relationship between the explanatory variables using a `pairs` plot. 

```{r,out.width="65%",fig.align='center', fig.alt="A scatterplot matrix of the three explanatory variables x1, x2 and x3 from the artificial data set. We see that x1 and x3 have a strong linear relationship, while x2 is shows no association with the other two explanatory variables." }
pairs(x1 ~ x2 + x3)
```

The scatterplot matrix illustrates the strong linear relationship between $X_1$ and $X_3$.

Now let us examine how this linear relationship affects the inference. First we consider a linear model of $Y$ on all three explanatory variables, that is $X_1$, $X_2$ and $X_3$. 

```{r}
m2 <- lm(y ~ x1 + x2 + x3)
summary(m2)
```
We note that the coefficient for $X_2$ is significant and its estimate is close to the true value 1. However, neither of the coefficients of $X_1$ and $X_3$ are significant, despite the fact that both predictor variables were used when we generated $Y$. Furthermore both have values far away from the true value of the corresponding parameter. In fact $X_3$ has a negative coefficient (albeit it not being significantly different from zero.)

Recall that, due to how we created the values for the explantory variables, we have that $X_3 \approx 2 X_1$. Now consider the estimated coefficients rounded to the nearest integer, Then,  using $X_3 \approx 2 X_1$, the mean function of the fitted model is roughly 
$$ 5X_1+X_2 - X_3 \; \approx \;  5 X_1+X_2 - 2 X_1 \; = \; X_1 + X_2 + 2 X_1 \; \approx \; X_1 + X_2 + X_3.$$ Thus the predictions from the model will be consistent with the model that generated the data.


Next consider a linear model of $Y$ on $X_2$ and $X_3$. 

```{r}
m1 <- lm(y ~ x1 + x2)
summary(m1)
```

We note that the coefficients for both $X_2$ and $X_3$ are significant. While the coefficient for $X_2$ is close to the true value of 1, the coefficient for $X_1$ is around 3. Taking into account again that $X_1 \approx 2 X_3$ we observe that the mean function is roughly given 
$$ 3X_1 +  X_2 \quad \approx \quad  X_1 +  X_2 + X_3.$$
Thus, while the existence of multi-collinearity creates with issues in identifying the effects of individual variables, it does not tend to affect prediction. 


We can quantify multi-collinearity using variance inflation factors. The $j$th variance inflation factor ${\rm VIF}_j$ is the multiplicative factor by which the variance of the estimator $\hat{\beta_j}$ is inflated due to (approximate) collinearity.

The variance inflation factors  be computed using `vif` from car package. As a rule of thumb we regard a ${\rm VIF}_j> 10$ as a concern.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(car)
df_vif <- round(vif(m2), 2)
kableExtra::kable(df_vif, col.names=c("Variable", "VIF"))
```

