---
title: "Practical Report 8"
output:
   pdf_document: 
     keep_tex: true
     fig_caption: yes
     highlight: monochrome
   html_document:
    df_print: paged
    code_folding: hide
fontsize: 12pt
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
require(knitr)
pdf_doc <- knitr::is_latex_output()
if (pdf_doc) {opts_chunk$set(echo=FALSE, fig.align="center", out.width="70%")} else {opts_chunk$set(echo=TRUE)}
```

# Polynomial regression

## Part A

In Practical 7 we observed a non-linear and non-monotonic relationship
between the explanatory variable and the response. We fit four
polynomial regression models of increasing complexity to the data.

```{r, fig.cap="A scatterplot of the data and the fitted curves."}
# Load and plot data 
poldata <- read.csv("../data/poldata.csv")
plot(y ~ x, data=poldata, pch=16)
# Fit the four polynomial models
m1 <- lm(y ~ x,                            data=poldata) #Linear model
m2 <- lm(y ~ x + I(x^2),                   data=poldata) #Quadratic model
m3 <- lm(y ~ x + I(x^2) + I(x^3),          data=poldata) #Cubic model
m4 <- lm(y ~ x + I(x^2) + I(x^3) + I(x^4), data=poldata) #Quartic model

# Vector of x values at which to compute  predicted y value
ax <- seq(from=2, to=6, length=101)
# Assign fitted models to a list
mlist <- list(m1, m2, m3, m4)
# Loop over list of models
for (i in seq_along(mlist)) {
    # Compute fitted values
    p <- predict(mlist[[i]], newdata=list(x=ax))
    # Plot smooth line through fitted values
    lines(ax, p, col=5-i, lwd=2, lty=i)
}
# Add a legend
legend("topright", lwd=2, col=4:1, legend=1:4, lty=1:4, title="degree")
```

We can see that the fitted cubic and quartic model are virtually
indistinguishable in the plot.

## Part B

To get the unbiased estimate for the error variance, we need the
residual standard error, which can be derived by `summary(m4)$sigma` or
just looking for it in the summary table.

```{r}
summary(m4)
print(paste0("The error variance is ", round(summary(m4)$sigma^2, 4)))
```

## Part C

Let the formula for the quartic model be represented algebraically as $$
Y \quad = \quad  \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \beta_4 x^4 + \epsilon.
$$ To test $H_0: \beta_4 = 0$ versus $H_1: \beta_4 \neq 0$ we use a
$t$-test. The number of units of observation in the dataset is $n=14$
and the quartic model has $p=5$ parameters. Hence, under $H_0$, the
$T$-statistic, defined as
$\frac{\widehat{\beta_4}}{{\rm std.error}(\widehat{\beta_4})}$ has a
$t$-distribution on $n-p=14-5 = 9$ degrees of freedom. As is listed in
the 4th row of the table of coefficients, the observed value of the
$T$-statistic is $T_{\rm obs} = -0.079$ which has a $p$-value of
$0.939 > 0.05$. Thus fail to reject the null hypothesis that
$\beta_4=0$. We conclude that the quartic model is not a significant
improvement over the cubic model. Hence we do not include a quartic term
in the model and instead retain the cubic model.

**Alternative:**

We can also compute the sequential ANOVA table for the quartic model.
This follows the hypothesis test given by the definition of the ANOVA
table.

```{r}
anova(m4)
```

## Part D

It is not possible to read from the model summary for the quartic model
whether a cubic polynomial is needed. The degree of freedom will be
different for different models. And the $p$-value reported for the cubic
term may change after removing the quartic term from the model.
**However, we can read from the ANOVA table for the hypothesis test
whether including the third order term.**

## Part E

Let the formula for the cubic model be represented algebraically as $$
Y = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \epsilon.
$$ We would like to test $H_0: \beta_3 = 0$ versus
$H_1: \beta_3 \neq 0$. The number of units of observation in the dataset
is $n=14$ and the cubic model has $p=4$ parameters. Hence, under $H_0$,
the $T$-statistic, defined as
$\frac{\widehat{\beta_3}}{{\rm std.error}(\widehat{\beta_3})}$ has a
$t$-distribution with $n-p=14-4 = 10$ degrees of freedom.

The observed value of the $T$-statistic is $$
T_{\rm obs}  \quad = \quad \frac{0.059}{0.009} \quad = \quad  6.56.
$$ The associated $p$-value is given by $$
p \quad = \quad  2 \Big[1- F_{t_{10}}(6.56)\Big].
$$ (We can compute $F_{t_{10}}(6.56)$ using the command
`pt(q=6.56, df=10)`.)

```{r}
t_obs <- summary(m3)$coefficients[4, 1] / summary(m3)$coefficients[4, 2]
print(paste0("The p-value is ", 2*(1-pt(q=t_obs, df=10))))
summary(m3)
```

# Multicollinearity

In the following we illustrate the issues that arise when we have multi-collinear predictors. We start by simulating realisations of 3 explanatory variables $X_1$, $X_2$ and $X_3$, such that $X_1$ and $X_3$ have a strong linear relationship. We then sample realisations of the response variable from the model
$$
Y_i \quad = \quad 1 + x_{i1} + x_{i2} + x{i_3} + \epsilon_i,
$$
where $i=1, \ldots, 100$ and $\epsilon_1, \ldots, \epsilon_{100}$ are iid N(0, $0.5^2$) distributed.

```{r}
# Producing realisations from the explanatory variables
set.seed(8)
x1 <- runif(100)
x2 <- runif(100)
x3 <- 2 * x1 + rnorm(100,0,0.02)
# Sampling response observations
y  <- 1 + x1 + x2 + x3 + rnorm(100,0,0.5)
```

We can illustrate the relationship between the explanatory variables using a `pairs` plot. We can see a strong linear relationship between $X_1$ and $X_3$.

```{r,out.width="65%",fig.align='center', fig.cap="A scatterplot matrix of the three explanatory variables x1, x2 and x3 from the artificial data set."}
pairs(x1 ~ x2 + x3)
```

First we consider a linear model of $Y$ on all three explanatory variables, that is $X_1$, $X_2$ and $X_3$.

```{r}
m2 <- lm(y ~ x1 + x2 + x3)
summary(m2)
```

We note that the coefficient for $X_2$ is significant and its estimate is close to the true value 1. However, neither of the coefficients of $X_1$ and $X_3$ are significant, despite the fact that both predictor variables were used when we generated $Y$. Furthermore both have values far away from the true value of the corresponding parameter. In fact $X_3$ has a negative coefficient (albeit it not being significantly different from zero.)

Recall that, due to how we created the values for the explantory variables, we have that $X_3 \approx 2 X_1$. Now consider the estimated coefficients rounded to the nearest integer, Then,  using $X_3 \approx 2 X_1$, the mean function of the fitted model is roughly 
$$
5X_1+X_2 - X_3 \; \approx \;  5 X_1+X_2 - 2 X_1 \; = \; X_1 + X_2 + 2 X_1 \; \approx \; X_1 + X_2 + X_3.
$$
Thus the predictions from the model will be consistent with the model that generated the data.

Next consider a linear model of $Y$ on $X_2$ and $X_3$.

```{r}
m1 <- lm(y ~ x1 + x2)
summary(m1)
```

We note that the coefficients for both $X_2$ and $X_3$ are significant. While the coefficient for $X_2$ is close to the true value of 1, the coefficient for $X_1$ is around 3. Taking into account again that $X_1 \approx 2 X_3$ we observe that the mean function is roughly given 
$$
3X_1 +  X_2 \quad \approx \quad  X_1 +  X_2 + X_3.
$$
Thus, while the existence of multi-collinearity creates with issues in identifying the effects of individual variables, it does not tend to affect prediction. 

We can quantify multi-collinearity using variance inflation factors. The $j$th variance inflation factor ${\rm VIF}_j$ is the multiplicative factor by which the variance of the estimator $\hat{\beta_j}$ is inflated due to (approximate) collinearity.

The variance inflation factors be computed using `vif` from car package. As a rule of thumb we regard a ${\rm VIF}_j> 10$ as a concern.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(car)
round(vif(m2), 2)
```
